Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-05-29 icass.Rnw
Revisions until 6.2.1

\subsection{Income}
Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1990 to 2005.

\subsection{Education}
Education is here analysed only for the period 1990 to 2005. Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden. It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.

@@

Models are fitted separately for each time period, from  1851 to
2013, with the exception of the period 1951--1960. We stratify on HISCLASS
and control for the covariates {\tt sex} and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
Figure~\ref{fig:cumsw}.

@@

Figure~\ref{fig:cumsw} shows mortality levels during the different periods. 
Obviously, and not surprisingly, mortality is decreasing over time, and we can 
already identify some interesting social patterns in mortality which will be 
commented more on below. Apart from
the social pattern that will be commented on below, there are a couple of
observations to be made from the figure. Women have consistently
better survival than men. This is a well-known phenomenon in Swedish
demographic history. Male mortality has almost always been higher than
female in all age groups \citep{wil99, suwi07}. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. The compression that is
visible for later decades is partly explained by the decreasing general
mortality. It is not
obvious that differences between social classes have diminished. This
conclusion has been confirmed by separate analyses (results not shown). There is no
clear indication of homogenisation of mortality levels between social
classes.

However, we are mainly interested in illustrating the change in relative 
position among the HISCLASSES over time rather than the absolute levels, so we 
\emph{standardise} the levels in each time period. It is illustrated in
Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
relative mortality levels for the different classes.

@@

Notice that we have not analysed the 1950s due to
the fact that we only have data for some parishes at that time,
consequently having no corresponding population for that
decade. Furthermore, we need to be cautious when interpreting the results for the 1960s in particular, but also for the 1970s due to much missing information on HISCLASS for these decades, something that have restricted the size of the population observed. The changing economy
also have consequences for the analysis, particularly when it comes to the
agricultural sector. While farmers were common occupations during the first
half of the 20th century, these occupations has constantly decreased in
numbers thereafter. This means that the second HISCLASS group has somewhat different character during the last periods in the figures.

The social pattern of mortality among women is quite much in accordance with previous
analyses from the present research group on other regions in northern
Sweden and for other periods \citep{edli11,edbr12,segb16}. The highest
social class usually have comparatively low mortality, while working
class women have high. This is especially the case during the latest
decades of the century and until 2013. We once again remind that the results from the 1960s must be interpreted with caution due to the large proportion of missing cases.

A different pattern appears among men. The groups we would expect to have
the lowest mortality, if we assume that access to different resources and
having a high status determines survival, instead have the highest during the nineteenth and
much of the 20th century. This is also in accordance with our previous studies
\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
low mortality until the 1960s. This changes however towards the end the 20th century. From
the 1980s, there is a clear advantage when it comes to survival for higher
managers and professionals. The expected social gradient in mortality has developed. This is also what we usually find in analyses
of social health inequalities in present-day Sweden as well as in most
other countries. In the regions we are studying, this is however a quite
recent phenomenon.

@@

We now turn to the elderly, i.e. those that mostly have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure.

The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level. Mortality among women was, as expected, lower than among
men. We find a clear decline in mortality, but this decline did not
set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the
Swedish development where we find a turning point at this time with better
survival among elderly, to a large extent caused by lower mortality in
cardiovascular diseases. The results for the 1960s are however uncertain as
discussed above, especially for women due to the difficulty to determine HISCLASS for the majority of them.

Concerning the social pattern, the pattern is a little more difficult to interpret, compared to the younger age group (the new analyses differ somewhat from the ones in \citet{segb16}). For women the results are mixed. We do not find the same clear advantage for elite women during period until 1950, and neither do we find any large disadvantage for working class elderly women. In some periods, the social differences were small, and the elite even had substantially higher mortality during the early 20th century. Notice however that this group was rather small at the time. During the recent decades, the pattern is not very clear, but for the two last periods, the elite have the lowest mortality. It is however not possible to identify a clear social gradient in mortality.

For men we find indications of the same intriguing change among elderly as for
the adult population 40--64 years. Mortality was in fact substantially
higher among men from the elite  but the disadvantage is mainly apparent during the 19th century. Differences are fairly small during the first half of the 20th century. Only during the second half of the 20th century, we find periods with much better survival for the highest social class while retired workers had fairly high mortality. But even in recent times, the pattern is not completely clear. During the last periods, we do not find a social gradient on HISCLASSES among elderly men (even in this case, the new analyses differ somewhat from the ones in \citet{segb16}). This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades. See Figure~\ref{fig:cumswold}.

@@

We now turn to a special analysis of mortality for the two dominating causes of death among the adult and elderly population. We start with cardiovascular diseases. Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group. Belonging to working class, having less education and low income increases the mortality risks. We would therefore expect to find a clear social pattern in the present analysis.

First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ. Thus, many deaths during the 19th century were not diagnosed at all, and there was still a substantial part of all death in the early twentieth century where death cause is designated as unknown or dying from old age, despite reporting causes of death had become mandatory. For the period 1901-1925, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, but at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent. Are there many cardiovascular deaths disguised there?
<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

As before, models are fitted separately for each ten-year time period, from  1851 to
2013. We stratify on HISCLASS
and use the covariates {\tt sex} and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

The authors of this paper have scrutinised different aspects of the social
pattern of mortality in northern Sweden for quite some time, focusing on
different regions, different age groups and partly different social
categories. Already the earliest findings, concerning the social inequality
in a rapidly growing town in an industrial district, surprised and at least
at first confused the author \citep{edv92}. The expected social gradient in
survival, meaning that people from higher social classes always had an
advantage, was in many cases, particularly among adult men, not found
despite the well-documented socio-economic differences in this
society. There were however large differences between men and women and for
different age groups. Further studies have confirmed these
findings \citep{edli11, edbr12}. These earlier studies have ended with an
assumption that distinct and consistent differences to the advantage of higher
classes developed somewhere during and probably quite early in the 20th
century. Until now we have however not been able to investigate the
continuous development on mortality differences among the adult and elderly
population due to no available data. This is the first time it has been
possible, and the results continue to surprise us. The consistent advantage
for higher social classes do not appear until quite late in the regions we
have analysed. The social pattern of female mortality are closer to what is
expected, but a social gradient in male adult mortality seems to be a
recent phenomenon. Furthermore, this transition and the class pattern of mortality differ depending on age group. Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly, a somewhat modified result from what we found in the first and preliminary analyses presented in Leuven this year \citep{segb16b}.

>>>>

\subsection{Income}
Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.

\subsection{Education}
Education is here analysed only for the period 1990 to 2013. Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden. It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.

@@

Models are fitted separately for each time period, from  1851 to
2013, with the exception of the period 1951--1960 and separately for sex. We stratify on HISCLASS
and control for the covariates {\tt   civst} and urban/rural residence. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

%The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
%Figure~\ref{fig:cumsw}.

@@

Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval 40--64 during the different periods. Obviously, and not surprisingly, mortality is decreasing over time for both sexes. Women have consistently better survival than men. This is a well-known phenomenon in Swedish
demographic history. Male mortality has almost always been higher than
female in all age groups \citep{wil99, suwi07}. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. The compression that is
visible for later decades is partly explained by the decreasing general
mortality and is not necessarily representing a homogenisation of relative risks.

By \emph{standardising e} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
relative mortality levels for the different classes.

@@

Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade. Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above. The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector. While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter. This means that the middle class has somewhat different character during the last periods in the figures.

The most striking result of the analysis is the different patterns for men and women. The social pattern of mortality among women is quite much in accordance with previous
analyses from the present research group on other regions in northern
Sweden and for other periods \citep{edli11,edbr12}. The highest
social class usually have comparatively low mortality, while working
class women have high. There is however no substantial change in the relative risks between the classes. Only the group with unknown social class is distinguished by moch higher mortality from the 1980s onwards. This can however have to do with the differences how social position for women is defined before and after 1950. After 1950 it is based on the own occupation. Being without occupation might include women that are more frail both when it comes to health and socially.

A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
\citep{edli11, edbr12}. Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class. The relative positions change radically towards the end the 20th century. From the 1980s, there is a clear advantage when it comes to survival for higher
managers and professionals. The relative differences are however more compressed than for the period until 1950. The expected social gradient in mortality has developed. This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
other countries. In the regions we are studying, this is however a quite
recent phenomenon. Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards. The assumption made for women of unknowns social class is probably even more relevant for men. They probably represent groups in society that are more vulnerable when it comes to mortality.

A further observation can be made from Figure~\ref{fig:cumsw}. The gender difference in death hazards is very big for the élite group from 1851--1950, men having about three times higher risks than women in the same class. In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.

@@

We now turn to the elderly, i.e. those that in most cases have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure.

The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history. Mortality among women was, as expected, lower than among men. We find a clear decline in mortality, but the fasted decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.

Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}). For women the results are mixed. Once again, too much missing data on social class leaves out the results for the 1960s and 1970s. We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time. In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century. Notice however that this group was rather small at the time. The social differences continued to be small during the last decades of the 20th century. Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.

For men we find indications of the same intriguing change among elderly as for
the adult male opulation 40--64 years. Mortality was in fact substantially
higher among élite men  but their disadvantage was primarily apparent during the 19th century. Differences are fairly small during the first half of the 20th century. Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison. ). This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.

@@

In this section, the focus is on analyses of mortality for the two dominating causes of death among the adult and elderly population. We start with cardiovascular diseases. Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group. Belonging to working class, having less education and low income increases the mortality risks. We would therefore expect to find a clear social pattern in the present analysis.

First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ. Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers. There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory. For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases. However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent. It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and the increase therefore is explained by improvements in how deaths were diagnosed.
<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013 and . We stratify on HISCLASS
and use the covariates urban/rural and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

The authors of this paper have scrutinised different aspects of the social
pattern of mortality in northern Sweden for quite some time, focusing on
different regions, different age groups and partly different social
categories. Already the earliest findings, concerning the social inequality
in a rapidly growing town in an industrial district, surprised and at least
at first confused the author \citep{edv92}. The expected social gradient in
survival, meaning that people from higher social classes always had an
advantage, was in many cases, particularly among adult men, not found
despite the well-documented socio-economic differences in this
society. There were however large differences between men and women and for
different age groups. Further studies have confirmed these
findings \citep{edli11, edbr12}. These earlier studies have ended with an
assumption that distinct and consistent differences to the advantage of higher
classes developed somewhere during and probably quite early in the 20th
century. Until now we have however not been able to investigate the
continuous development on mortality differences among the adult and elderly
population due to no available data. This is the first time it has been
possible, and the results continue to surprise us. The consistent advantage
for higher social classes do not appear until quite late in the regions we
have analysed. The social pattern of female mortality are closer to what is
expected, but a social gradient in male adult mortality seems to be a
recent phenomenon. Furthermore, this transition and the class pattern of mortality differ depending on age group. Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.
